---
title: "Untitled"
author: "RN7"
date: "February 24, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Bullet Charts for Monitoring KPIs

This is the repo for the article found [here]() on creating bullet charts for monitoring KPIs over time. The article mainly focused on `M&E` deliverables, which are called **"Indicators"**, but this post and repo will also come in handy for anyone that depends on Key Performance Indicators (KPI)s or those that need to track progress against a target.

Let's look at the kind of data frame you would need to create the chart:

```{r data, echo=FALSE}

suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(ggrepel))

## Inputs
forYear = 2018 ## Specify Year the analysis represents
FY = TRUE       ## Is this a fiscal year? (as opposed to calendar year)
ProjectStartDate <- "2016/03/01"

df <- structure(list(IndicatorName = c("Ind 04", "Ind 05", "Ind 07", 
                                       "Ind 11", "Ind 17", "Ind 18"), 
                     Actual = c(3, 437, 20, 44, 1,10000), 
                     Actual_lastWeek = c(3, 420, 18, 20, 1, 10000), 
                     Actual_lastYear = c(3, 50, 20, 2000, 1, 10000), 
                     Target = c(14, 81, 21, 10327, 5, 20000)), 
                row.names = c(NA, -6L), class = "data.frame", 
                .Names = c("IndicatorName", "Actual", "Actual_lastWeek", "Actual_lastYear", "Target"))

## If df is empty, break function and just output an empty chart
if (nrow(df) == 0){
  OutputPlot <- "No data yet"
  return()
}

## Protect against the case where there are no actuals:
if (is.null(length(df$Active))) df$Actual <- 0

## Take percents, but protecting against there being no targets
df$Perc <-     df$Actual/(df$Target + 0.0000000000001) * 100
df$PercWeek <- df$Actual_lastWeek/(df$Target + 0.0000000000001) * 100
df$PercYear <- df$Actual_lastYear/(df$Target + 0.0000000000001) * 100

## But truncate results if any are greater than 100
df$Perc[df$Perc > 100] <- 100
df$PercWeek[df$PercWeek > 100] <- 100
df$PercYear[df$PercYear > 100] <- 100

#### Figure out today, within the Fiscal Year 
## If desired, force the project to respect the startdate
if (FY == TRUE){
  ## Control for the possibility that project starts in a month other than October
  if(month(ProjectStartDate) != 10) {ProjectStartDate <-  as.Date(paste("01 10", year(ProjectStartDate)), format = "%d %m %Y")}
  ## Grab Percent of year
  PercentTime<- 
    as.numeric((Sys.Date() - as.Date(paste(format(as.Date(ProjectStartDate), "%d %b"),
                                           forYear - 1), format = "%d %b %Y"))/365.25*100)
} else {
  ## Grab Percent of year
  PercentTime<- 
    as.numeric(Sys.Date() - as.Date(ProjectStartDate))/365.25*100
}

## Truncate Percent line if greater than 100 (for example, doing the analysis after year end)
if(PercentTime > 100) PercentTime <- 100

## Calculate how far behind TODAY the Indicator Percent is
df$BehindBy <- df$Perc - PercentTime

## Calculate how far behind TODAY (VALUE) for text, but remove NAs
df <- df %>%
  mutate(text = PercentTime/100 * df$Target - df$Actual)
df$text[df$BehindBy > 0] <-"OK!"
df$text[df$BehindBy <= 0 & !is.na(df$BehindBy)] <- 
  paste("Need ", round(as.numeric(df$text[df$BehindBy <= 0  & !is.na(df$BehindBy)])), " more", sep = "")

          ## Calculate behind from last week
          df$BehindFromLastWeek <- df$Actual - df$Actual_lastWeek
          
          ## Truncate BEHIND BY at 0, LowLevel, and Percent to 100
          #### Set "low level" here. Low level is the Value that we consider tolerable to be behind the indicators by
          LowLVL <- -.20 * PercentTime
          df$BehindBy[df$BehindBy > 0] <- 0
          df$BehindBy[df$BehindBy < LowLVL] <- LowLVL

```


```{r}
glimpse(df)
```

As we can see...

The variables for the indicators/KPIs are as follows:

* `Actual`: the value of the indicator at the current time of viewing ("Today")
* `Actual_lastWeek`: Last week's value of the indicator
* `Actual_lastYear`: Last year's value of the indicator
* `Target`: the target value for the indicator (used to calculate the percent variables)

Percentages are calculated by:

```{r}
## Take percents, but protecting against there being no targets
df$Perc <-     df$Actual/(df$Target + 0.0000000000001) * 100
df$PercWeek <- df$Actual_lastWeek/(df$Target + 0.0000000000001) * 100
df$PercYear <- df$Actual_lastYear/(df$Target + 0.0000000000001) * 100

## But truncate results if any are greater than 100
df$Perc[df$Perc > 100] <- 100
df$PercWeek[df$PercWeek > 100] <- 100
df$PercYear[df$PercYear > 100] <- 100

```

* `Perc`: Value of indicator as percent of yearly taget and percent of the year at the current time
* `PercWeek`: Last week's value of the indicator as percent of yearly target and percent of the year
* `PercYear`: Last year's value of the indicator as percent of yearly target and percent of the year

`BehindBy` is calculated by: `df$Perc - PercentTime` and shows how far behind the current value of the indicator is to the target value for the current time

The text can be shown by calculating:

```{r}

## Calculate how far behind TODAY (VALUE) for text, but remove NAs
df <- df %>%
  mutate(text = PercentTime/100 * df$Target - df$Actual)
df$text[df$BehindBy > 0] <-"OK!"
df$text[df$BehindBy <= 0 & !is.na(df$BehindBy)] <- 
  paste("Need ", round(as.numeric(df$text[df$BehindBy <= 0  & !is.na(df$BehindBy)])), " more", sep = "")

```

* `text` = "OK!": Shows that the current value of the indicator meets the target value for the current time
* `text` = "Need __ more": Shows exactly how much more of the indicator is needed to reach the target value for the current time


Here's one of the versions of the graph!

```{r plot, warning=FALSE, echo=FALSE}

ggplot(df, aes(IndicatorName)) + 
  geom_col(aes(y = Perc, width = 0.1, fill = BehindBy), color = "black") + 
  scale_fill_gradient("Indicator\nBehind By:", limits = c(LowLVL, 0), low = "red", high = "green",
                      labels = c("On Time", "Slightly Behind", "Behind Schedule", ""),
                      guide = FALSE) +
  geom_point(aes(y = PercWeek, shape = "Last Week"), size = 6, stroke = 1) +
  geom_point(aes(y = PercYear, shape = "Last Year"), size = 6, stroke = 1) +         
  scale_shape_manual(" ", values = c(23, 21)) +
  geom_col(aes(y = 100, width = 0.5), alpha = 0.25) +
  geom_text(y = 1, aes(label = text), vjust = -1, hjust = 0) +
  geom_hline(yintercept = PercentTime, alpha = 0.33) +
  annotate("text", x = 0, y = PercentTime + 1.5, hjust = 0, label = "Today", 
           angle = 90, alpha = 0.5, size = 5) +
  coord_flip() +  
  labs(y = "Percent of Yearly Target\n&\n Percent of Year",
       x = " ") +
  ggtitle(paste("Ongoing Indicator Accomplishment (", forYear, ")", sep = "")) +
  theme_minimal() + 
  theme(axis.text.y = element_text(size = 15, face = "bold"),
        axis.title.x = element_text(face = "bold", size = 10),
        axis.text.x = element_text(face = "bold", size = 14),
        title = element_text(face = "bold", size = 14),
        plot.title = element_text(hjust = 0.5), 
        plot.subtitle = element_text(hjust = 0.5, size = 8),
        legend.position = c(0.8, -0.12),
        legend.key.size = unit(1.5, "lines")) +
  expand_limits(x = 6.75, y = 102)

```

For something more similar to Stephen Few's chart there is also this version:

```{r multiple_bars, echo=FALSE, warning=FALSE}

ggplot(df, aes(IndicatorName)) + 
  geom_col(aes(y = PercWeek, width = 0.5), alpha = 0.6) +
  geom_col(aes(y = PercYear, width = 0.75), alpha = 0.3) +
  geom_col(aes(y = Perc, width = 0.15, fill = BehindBy), color = "black") + 
  scale_fill_gradient("Indicator\nBehind By:", limits = c(LowLVL, 0), low = "red", high = "green") +
  geom_text(y = 1, aes(label = text), vjust = -1.5, hjust = 0, size = 4) +
  geom_hline(yintercept = PercentTime, alpha = 0.33) +
  annotate("text", x = 0, y = PercentTime + 1.5, hjust = 0, label = "Today", angle = 90, alpha = 0.5, size = 5) +
  coord_flip() +
  labs(y = "Percent of Yearly Target\n&\n Percent of Year",
       x = " ") +
  ggtitle(paste("Ongoing Indicator Accomplishment (", forYear, ")", sep = "")) +
  theme_minimal() + 
  theme(axis.text.y = element_text(size = 15, face = "bold"),
        axis.title.x = element_text(face = "bold", size = 10),
        axis.text.x = element_text(face = "bold", size = 12),
        title = element_text(face = "bold"),
        plot.title = element_text(hjust = 0.5), 
        plot.subtitle = element_text(hjust = 0.5, size = 8),
        legend.position = "none")

```







